Well-quasi-orderings on word languages

### 2024-08-15 src/metadata.tex
typos

\title{Well-quasi-orderings on word languages}
\author{%
        Roland Guttenberg \and
        Nathan Lhote \and
        Aliaume Lopez \and
        Corto Mascle \and
        Vincent Michielini \and
        Lia Schütze \and
        Omid Yaghoubi}

\newcommand{\makeabstract}{
\begin{abstract}
    The set of finite words over a well-quasi-ordered set is itself
    well-quasi-ordered. This seminal result by Higman is a cornerstone
    of the theory of well-quasi-orderings and has found numerous
    applications in computer science. However, this result is based on a
    specific choice of ordering on words, the (scattered) subword
    ordering. In this paper, to what extent other natural orderings
    (prefix, suffix, infix) on words can be used to derive Higman-like
    theorems. More specifically, we are interested in characterizing
    \emph{languages} of words that are well-quasi-ordered under these
    orderings. We show that a simple characterization is possible for
    the prefix and suffix orderings, and that under extra regularity
    assumptions, this also lifts to the infix ordering.
\end{abstract}
}

>>>>

\title{Well-quasi-orderings on word languages}
\author{%
        Roland Guttenberg
    \thanks{Technical University of Munich, Germany}
     \and
        Nathan Lhote
    \thanks{Aix-Marseille University}
     \and
        Aliaume Lopez
    \thanks{University of Warsaw}
     \and
        Corto Mascle
    \thanks{LaBRI, University of Bordeaux}
     \and
        Vincent Michielini
    \thanks{University of Warsaw}
     \and
        Lia Schütze
    \thanks{MPI-SWS, Germany}
     \and
        Omid Yaghoubi
    \thanks{University of Warsaw}
    }

\newcommand{\makeabstract}{
\begin{abstract}
    The set of finite words over a well-quasi-ordered set is itself
    well-quasi-ordered. This seminal result by Higman is a cornerstone
    of the theory of well-quasi-orderings and has found numerous
    applications in computer science. However, this result is based on a
    specific choice of ordering on words, the (scattered) subword
    ordering. In this paper, we describe to what extent other natural
    orderings (prefix, suffix, infix) on words can be used to derive
    Higman-like theorems. More specifically, we are interested in
    characterizing \emph{languages} of words that are well-quasi-ordered
    under these orderings. We show that a simple characterization is
    possible for the prefix and suffix orderings, and that under extra
    regularity assumptions, this also lifts to the infix ordering.
\end{abstract}
}

### 2024-08-15 src/prefixes.tex
typos + amalgamation + conclusion
typos

The next remark that we make is that $\Sigma^*$ is not \kl{well-quasi-ordered}
by the \kl{prefix relation} as soon as $\Sigma$ contains two distinct letters
$a$ and $b$. As an example of infinite \kl{antichain}, we can consider the set
of words $a^n b$ for $n \in \Nat$. As mentioned in the introduction, there are
some languages however that are \kl{well-quasi-ordered} by the \kl{prefix
relation}. A simple example being the (regular) language $a^* \subseteq
\set{a,b}^*$, which is order-isomorphic to natural numbers with their usual
orderings $(\Nat, \leq)$.

@@

\begin{figure}
    \centering
    \caption{The tree of prefixes over the alphabet $\set{a,b}$,
        with an \kl{antichain} in red, and
        an \kl{antichain branch} in blue.}
\end{figure}

@@

\begin{definition}
    An \intro{antichain branch} for a language $L$ is an infinite 
    branch $B$ of the \kl{tree of prefixes} such that from every point of the branch, 
    one can reach a word of $L$ that is not in the branch.
\end{definition}

Let us illustrate the notion of \kl{antichain branch} over the alphabet $\Sigma
= \set{a,b}$, and the language $L = a^* b$. In this case, the set $a^*$ (which
is a branch of the \kl{tree of prefixes}) is an \kl{antichain branch} for $L$.
This holds because for any $a^k$ in the branch, $a \prefleq a^kb$ which is in
$L$ and not in the branch. In general, the existence of an \kl{antichain
branch} for a language $L$ implies that $L$ contains an infinite
\kl{antichain}, and because the alphabet $\Sigma$ is assumed to be finite, one
can leverage the fact that the \kl{tree of prefixes} is finitely branching to
prove that the converse holds as well.

@@

For the proof of the following theorem, we will use special notations to
describe the \intro{upwards closure} of a set $S$ for a relation $\preceq$,
which is defined as $\upset[\preceq]{S} \defined \setof{y \in \Sigma^*}{
\exists x \in S. x \preceq y}$. Anticipating the use of the symmetric notion of
\intro{downwards closure}, let us introduce its notation 
straight away as follows: $\dwset[\preceq]{S} \defined \setof{y
\in \Sigma^*}{ \exists x \in S. y \preceq x}$. Abusing notations, we will
write $\upset{w}$ and $\dwset{w}$ for the upwards and downwards closure of a
single element $w$, and omit the ordering relation when it is clear from the
context.

@@

As an immediate consequence, we have a very fine-grained understanding of the
\kl{ordinal invariants} of such \kl{well-quasi-ordered} languages, which can be
leveraged in bounding the complexity of algorithms working on such languages.

>>>>

The next remark we make is that $\Sigma^*$ is not \kl{well-quasi-ordered} by
the \kl{prefix relation} as soon as $\Sigma$ contains two distinct letters $a$
and $b$. As an example of infinite \kl{antichain}, we can consider the set of
words $a^n b$ for $n \in \Nat$. As mentioned in the introduction, there are
however some languages  that are \kl{well-quasi-ordered} by the \kl{prefix
relation}. A simple example being the (regular) language $a^* \subseteq
\set{a,b}^*$, which is order-isomorphic to natural numbers with their usual
orderings $(\Nat, \leq)$.

@@

@@

\begin{definition}
    An \intro{antichain branch} for a language $L$ is an infinite 
    branch $B$ of the \kl{tree of prefixes} such that from every point of the branch, 
    one can reach a word in $L \setminus B$.
\end{definition}

Let us illustrate the notion of \kl{antichain branch} over the alphabet $\Sigma
= \set{a,b}$, and the language $L = a^* b$. In this case, the set $a^*$ (which
is a branch of the \kl{tree of prefixes}) is an \kl{antichain branch} for $L$.
This holds because for any $a^k$, the word $a \prefleq a^kb$ belongs to $L
\setminus a^*$. In general, the existence of an \kl{antichain branch} for a
language $L$ implies that $L$ contains an infinite \kl{antichain}, and because
the alphabet $\Sigma$ is assumed to be finite, one can leverage the fact that
the \kl{tree of prefixes} is finitely branching to prove that the converse
holds as well.

@@

For the proof of the following theorem, we will use special notations to
describe the \intro{upwards closure} of a set $S$ for a relation $\preceq$,
which is defined as $\upset[\preceq]{S} \defined \setof{y \in \Sigma^*}{
\exists x \in S. x \preceq y}$. Anticipating the use of the symmetric notion of
\intro{downwards closure}, let us introduce it as follows: $\dwset[\preceq]{S} \defined \setof{y
\in \Sigma^*}{ \exists x \in S. y \preceq x}$. Abusing notations, we will
write $\upset{w}$ and $\dwset{w}$ for the upwards and downwards closure of a
single element $w$, omitting the ordering relation when it is clear from the
context.

@@

As an immediate consequence, we have a very fine-grained understanding of the
\kl{ordinal invariants} of such \kl{well-quasi-ordered} languages, which could be
leveraged in bounding the complexity of algorithms working on such languages.
